---
title: "tutorial_notes_part2"
author: "Amy R Hodgson"
date: "01/06/2019"
output: html_document
---

## How to customise reactions

### What is reactivity?

Example: entering values in cells in excel and manipulating them with fomulas to get result in different cells. Changing values will change result of formulas being applied (output). In shiny, changes in input objects propagate to output objects. Inputs are sliders/menus etc for user, output is tables/graphs etc. However, in shiny it can get more complex than this. Eg change in input could trigger other code to run on the serverside, or could create secondary reactive objects, and changes in these can trigger changes in output.

### Begin with reactive values

We can create an input object with an input function, eg `sliderInput(inputId = "num")`, where the inputId creates a value saved as `input$num` that can later be used in the server function. This value changes as user changes input to slider. Reactive values don't work on their own, and work together with a reactive function. You can't call a reactive value from outside of one. Example: `renderplot({ hist(rnorm(100, input$num))} )`. A reactive function is one that expects to take reactive values and knows what to do with them.

Important point that reactive values and reactive functions don't work without each other- common source of errors.

Can think of reactivity as a 2-step process. First step is `input$num` and second `output$hist <- renderPlot({ hist(rnorm(input$num)) })}`. `renderPlot` uses `input$num` to create histogram and save in `output$hist`. Can call object created by the reactive function the 'observer'. Together the observer and reactive value create the reactivity in the app.

### Reactivity in R is a 2-step process

1. Reactive values notify things downstream of the change, i.e. the reactive functions, when they become invalid (are changed by user).

2. The objects created by reactive functions respond. When have render function, it creates an object which re-runs the code to be passed to the render function to update itself. Ih this example, code then calls `input$num` and gets new input to create new graph. Can customise response by using correct reactive function.

### Recap: Reactive values

- Reactive values act as a data stream that flows through your apps. They are the changes that cause reactions downstream.

- The `input$` list is a list of reactive values. The values show the current state of the inputs.

- You can only call a reactive value from a function designed to work with them- i.e. a reactive function.

- Reactivity in R has 2 steps. Reactive values notify things downstream that they have changed, and the objects created by reactive functions respond. Process called invalidating.

### Reactive toolkit: 7 indespensible functions

Reactive functions:

1. Use a code chunk to build (and rebuild) an object. To understand more deeply, must ask what code will the function use to build the object?

2. The object will respond to changes in a set of reactive values. Think about which reactive values will the object respond to?

#### Display output with render*()

Family of functions that create a certain type of output. Render functions build output to display in the app, e.g. `renderPlot`. Takes one argument, which is a block of code {in curly braces} that the function uses to build and rebuild object. The object will respond to every reactive value in the code.

When notified that it is invalid, the object created by a render* function will re-run the entire block of code associated with it.

```{r, eval=FALSE}

library(shiny)

ui <- fluidPage(
  sliderInput(inputId = "num",
              label = "Choose a number",
              value = 25, min = 1, max = 100),
  textInput(inputId = "title",
            label = "Write a title",
            value = "Histogram of random normal values"),
  plotOutput("hist")
)

server <- function(input, output) {
  output$hist <- renderPlot( {
    hist(rnorm(input$num), main = input$title)
  })
}

shinyApp(ui = ui, server = server)

```

In this example, both reactive values `input$num` and `input$title` are used by render function to make plot. Changing either invalidates `output$hist`, which will then re-run code in render function to update itself.

#### Modularise code with reactive()

A slightly more complex example:

```{r, eval = F}
ui <- fluidPage(
  sliderInput(inputId = "num",
              label = "Choose a number",
              value = 25, min = 1, max = 100),
  plotOutput("hist"),
  verbatimTextOutput("stats")
)

server <- function(input, output) {
  output$hist <- renderPlot( {
    hist(rnorm(input$num))
  })
  output$stats <- renderPrint( {
    summary(rnorm(input$num))  
  })
}

shinyApp(ui = ui, server = server)
```

In this case, there is one reactive value `input$num` and two observer objects downstream, `output$hist` and `output$stats`, which update when input is invalidated. However, each object is generating its own histogram to describe- have two different datasets. Really want to generate and describe one dataset. 

Solution could be: instead of calling rnorm twice, call it once and save the dataset it creates, then use that dataset downstream where it is needed- i.e. `data <- ? rnorm(input$num)`. But how to save that dataset? Input contains reactive values, so need to use reactive function to work alongside it. Want it to be able to update when invalidated, communicate this to output.

Can do this with the function `reactive()`. This builds a reactive object (reactive expression in shiny vocabulary). Takes input objects and assembles them into some kind of output- could be eg df, value, plot. Makes the output available to use later in other code. 

Example: `data <- reactive( {rnorm(input$num) })`. Pass code block as argument, which reactive will use to build and rebuild object. The object will then respond to any reactive value in the code that changes.

A reactive expression is special in two ways:

1. Remember that it is a function, and so a reactive expression must be called as a function (with no arguments) e.g. `data()`. 

```{r, eval = F}
ui <- fluidPage(
  sliderInput(inputId = "num",
              label = "Choose a number",
              value = 25, min = 1, max = 100),
  plotOutput("hist"),
  verbatimTextOutput("stats")
)

server <- function(input, output) {
  data <- reactive({
    rnorm(input$num)
  })
  output$hist <- renderPlot( {
    hist(data())
  })
  output$stats <- renderPrint( {
    summary(data())  
  })
}

shinyApp(ui = ui, server = server)

```

Now if were to change input, `input$num` woud notify the reactive expression that it is invalid. The reactive expression would then notify anything downstream (eg hist) that they are also invalid. This then re-runs the render function, calling data from reactive expression. At this point the reactive expression re-runs its code, gets value from input$num. Reactive expression computes and saves its result and passes downstream. Then the second observer (eg stats) re-runs its code and calls reactive expression. As reactive expression saves result and knows when it is invalid or valid, it can simply pass saved (cached) value from when called previously by hist. If it knows it's invalid, it would re-run.

2. Reactive expressions cache their values (the expression will return its most recent value, unless it has become invalidated)

So:

- reactive() makes an object to use (in downstream code)
- Reactive expressions are themselves reactive. Can use them to modularise your apps.
- Remember to call a reactive expression like a function
- Reactive expressions cache their values to avoid unnecesary computation

#### Prevent reactions with isolate()

Going back to a previous example. When change slider or name, histogram updates. However, when changing the name, app may react faster than desired- e.g. histogram may refresh when halfway through typing new name. Want to stop app responding to title field to give time to enter full title before updating.

Can use isolate(), which is analogous to reactive(), but creates output that is not reactive. It returns the result as a non-reactive value. Example: `isolate( {rnorm(input$num) })`. Takes a block of code used to build the object, but the object will not respond to any reactive value in the code, i.e. changes to `input$title`. However if something else within the same function is invalidated, it will re-run code and changes to `input$title` will be made.

```{r, eval = F}
ui <- fluidPage(
  sliderInput(inputId = "num",
              label = "Choose a number",
              value = 25, min = 1, max = 100),
  textInput(inputId = "title",
            label = "Write a title",
            value = "Histogram of random normal values"),
  plotOutput("hist")
)

server <- function(input, output) {
  output$hist <- renderPlot( {
    hist(rnorm(input$num), 
         main = isolate({input$title})) # fetches value of title and gives back in non-reactive state
  })
}

shinyApp(ui = ui, server = server)
```

So summary:

- isolate() makes a non-reactive object
- use isolate() to treat reactive values like normal R values

#### Trigger code with observeEvent()

Going to learn to trigger code on the serverside based on changes in input objects, can include eg creating/downloading file. To do this going to use new type of input object: action button. Example:

`actionButton(inputId = "go", label = "Click me!")`

Takes an inputId for internal use (like *input functions) and a label to display. Goes in fluidPage(). This code would create a button saying Click me with no functionality (as nothing in server function).

The way to trigger code is with observeEvent(). This triggers code to run on server.

`observeEvent(input$clicks, {print(input$clicks)})`

Has different syntax to previous functions. First argument is either a reactive value or a vector of reactive values. These values are tracked by observeEvent and it responds as necessary. The observer invalidates only when the reactive value changes. It creates an observer object that exists in memory. The second argument is a block of code to run whenever the observer is invalidated. This clock of code can contain reactive values, but changes in these reactive values will not trigger observeEvent to invalidate. It is as if the block of code is wrapped in an isolate() function. This means you don't have to pair reactive values here with a reactive function as R will treat them as normal R values.

```{r, eval = F}

ui <- fluidPage(
  actionButton(inputId = "clicks",
               label = "Click me")
)

server <- function(input, output) {
  observeEvent(input$clicks, {   # don't save input, but it will be in app memory. Would need render function to display if desired
    print(as.numeric(input$clicks)) 
  })
}

shinyApp(ui = ui, server = server)

```

When run apps locally, your R session is the server for the app, so result is displayed in R console. Value of action button isn't meant to be meaningful though, simply to make a change that things can respond to.

There is also function called `observe()`. Parallel to observeEvent(). It also triggers code to run on server, and uses same syntax as render*(), reactive() and isolate (). 

`observe({ print(input$clicks) })`

Just give it single block of code to run. Then observe() will re-run that block of code any time the reactive values in the block of code changes. It will respond to every reactive value in the code. However observeEvent can be simpler to understand and more useful.

So:

- observeEVent() triggers code to run on the server
- allows to specify precidely which reactive values should invalidate the observer
- can use observe() for more implicit syntax, but tracks every reactive value in the code given to it

#### Delay reactions with eventReactive()

Sometimes just want to delay reactions of app rather than halt them. So using previous exmaple with histogram, but instead want to have an update button to trigger chanes to the graph rather than the slider. Can use eventReactive(), which makes a reactive expression that only responds to specific values. 

`data <- eventReactive(input$go, { rnorm(input$num) })`

Has similar syntax to observeEvent(). First argument is a reactive value or vector of reactive values to respond to, and the expression invalidates only when this value changes. Second argument is code that it will use to build and rebuild the object when it is invalidated. However, like with observeEvent(), it will treat reactive values in this code as if they have been isolated.

```{r, eval = F}

ui <- fluidPage(
  sliderInput(inputId = "num", 
    label = "Choose a number", 
    value = 25, min = 1, max = 100),
  actionButton(inputId = "go",
               label = "Update"),
  plotOutput("hist")
)

server <- function(input, output) {
  data <- eventReactive(input$go, {
    rnorm(input$num)
  })
  output$hist <- renderPlot({
    hist(data())
  })
}

shinyApp(ui = ui, server = server)

```

Now whenever `input$go` is invalidated, triggers update of plot downstream as observer it notified it is also invalid.

So:

- Use eventReactive() to delay reactions
- It creates a reactive expression, so you must call the results as a function
- You can specify precisely which reactive values should invalidate the expression

#### Manage state with reactiveValues()

Input values change whenever a user changes the input, and you cannot set these values in your code. However, reactiveValues() allows you to create your own list of reactive values which can be manipulated programmatically.

`rv <- reactiveValues(data = rnorm(100))`

It doesn't require any arguments, but you can give it arguments to populate the list created. So this creates a reactiveValues list that begins with one element named data, and that element will contain 100 rnorm values.

Below, have a reactive value with one element 'data'. Downstream is a histogram which takes dependency on `rv$data`. Whenever norm or unif are changed (clicked), it overwrites the value of `rv$data` and the render function will re-run its code, fetching new value of `rv$data` and updating plot.

```{r, eval = F}

ui <- fluidPage(
  actionButton(inputId = "norm", label = "Normal"),
  actionButton(inputId = "unif", label = "uniform"),
  plotOutput("hist")
)

server <- function(input, output) {
  
  rv <- reactiveValues(data = rnorm(100))
  
  observeEvent(input$norm, { rv$data <- rnorm(100)})
  observeEvent(input$unif, { rv$data <- runif(100)})
  
  output$hist <- renderPlot({
    hist(rv$data)
  })
}

shinyApp(ui = ui, server = server)

```

Using observeEvent, can overwite rv$data each time either button is clicked.

So:

- reactiveValues() creates a list of reactive vaues
- You can manipulate these values (usually with observeEvent)

Tips:

Keep in mind the end result of what app is going to be. Number of users, how hosted etc affects how should write code. Code may be repeated multiple times. If you write code outside of the server function, it will be run once every time a new R session is created. If you only need to run code once to set things up, eg ui and library, place this outside of the server function. 

Code inside the server function will be run once per user end user who logs into R session. There will be more users than R sessions, so code appearing in the server function will be run more often than code outside it- this takes up more resources and makes app slower. 

The code inside a render function will be run once per reaction, potentially many times. So should place bare minimum code inside a render function that is necessary for updating the output.

Secondly, if want to understand how R implements reactivity with shiny, can look through the source code for shiny, or read article about it.